Implementing Distributed Systems Directly\\with I/O Automata in ioa++

### 2011-09-30 papers/paper1/case_study.tex
Another pass.

A node in the ring passively waits for a connection from its successor and actively tries to connect to its successor.
The tcp\_ring\_automaton listed in examples/tcp\_lcr.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}.} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and channel to the successor.
The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
The tcp\_ring\_automaton uses timeouts and the techniques listed in Section~\ref{constellations} to manage the dynamic constellation of automata and recover from errors.
The tcp\_ring\_automaton contains two external actions:  one for receiving messages from a predecessor and one for sending messages to a successor.

>>>>

A node in the ring passively waits for a connection from its successor and actively tries to connect to its successor.
The tcp\_ring\_automaton listed in examples/tcp\_ring\_automaton.hpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_ring_automaton.hpp}.} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and the channel to the successor.
The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
The tcp\_ring\_automaton uses timeouts and the techniques listed in Section~\ref{constellations} to manage the dynamic constellation of automata and recover from errors.
The tcp\_ring\_automaton contains two external actions:  one for receiving messages from a predecessor and one for sending messages to a successor.

### 2011-09-30 papers/paper1/design.tex
Another pass.

\paragraph*{Actions of no-longer-existing automata}
Every action in the set of actions considered by the controller is scheduled by its associated automaton.
Since automata can be destroyed, an action in the set owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)}, assuming that automata turn over less frequently overall than the set of actions in the scheduler.

@@

\paragraph*{Automaton and Binding managers}
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating more complicated ones.
Automaton Managers and Binding Managers provide individualized facades to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be monitored using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created: thus, Binding Managers observe Automaton Managers.

\paragraph*{Generators}
A \emph{generator} is a single-use factory for creating an instance of an automaton.
Generators are used to produce new automata as part of the create system action and are visible as arguments in the various layers of the framework.

>>>>

\paragraph*{Actions of no-longer-existing automata}
Every action in the set of actions considered by the controller is scheduled by its associated automaton.
Since automata can be destroyed, an action in the set owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assigns each automaton a unique identifier assuming that the set of automata turn over less frequently overall than the set of actions in the scheduler.

@@

\paragraph*{Automaton and Binding managers}
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple constellations but becomes unwieldy when creating more complicated ones.
Automaton Managers and Binding Managers provide individualized facades to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be monitored using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created: thus, Binding Managers observe Automaton Managers.

\ifjournal
\paragraph*{Generators}
A \emph{generator} is a single-use factory for creating an instance of an automaton.
Generators are used to produce new automata as part of the create system action and are visible as arguments in the various layers of the framework.
\fi

### 2011-09-30 papers/paper1/evaluation.tex
Another pass.

This experiment indicates that concurrent execution is possible with ioa++ but a speed-up will only be achieved if (1) enough independent actions are enabled and (2) the duration of the independent actions overcome the overhead of ioa++.
The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the ioa::schedule call.
From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4316ns $\pm$ 3.5ns for the single threaded experiments and 6067ns $\pm$ 3.6ns for the multi-threaded experiments.
Similary, we calculated the scheduling overhead and found it to be 247ns $\pm$ 0.678ns for the single threaded experiments and 307ns $\pm$ 0.357ns for the multi-threaded experiments.
Both of these potentially can be improved with better scheduler design.
The synchronization time depends on the interactions among the system being executed, the scheduler, and the locking scheme used by the framework.
We plan to address scheduling and synchronization in future work.

>>>>

This experiment indicates that concurrent execution is possible with ioa++ but a speed-up will only be achieved if (1) enough independent actions are enabled and (2) the duration of the independent actions overcome the overhead of ioa++.
The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the ioa::schedule call.
From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4316ns $\pm$ 3.5ns for the single threaded experiments and 6067ns $\pm$ 3.6ns for the multi-threaded experiments.
Similarly, we calculated the scheduling overhead and found it to be 247ns $\pm$ 0.678ns for the single threaded experiments and 307ns $\pm$ 0.357ns for the multi-threaded experiments.
Both of these potentially can be improved with better scheduler design.
The synchronization time depends on the interactions among the automata comprising the system, the scheduler, and the locking scheme used by the framework.
We plan to address scheduling and synchronization in future work.

### 2011-09-30 papers/paper1/programming_model.tex
Another pass.

All automata inherit from the common base class ioa::automaton which implements the system action interface.
\ifjournal
The following code declares the asynch\_lcr\_automaton:
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
The asynch\_lcr\_automaton is declared as a C++ template where the UID template parameter must model the \emph{unique identifier} concept.
The generic programming facilities of the C++ language aid in the development of general purpose reusable components.
\fi

@@

A number of wrappers and macros are provided by ioa++ for declaring actions.
%%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
Local actions are declared using three functions corresponding respectively to precondition, effect, and scheduling.
The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
\ifjournal
For example, the send action of the asynch\_lcr\_automaton is declared:
\begin{lstlisting}
private:
  bool send_precondition () const { ... }
  UID send_effect () { ... }
  void send_schedule () const { ... }
public:
  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
\end{lstlisting}
\fi
Input actions are declared similarly except for the absence of a precondition.
The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
The macros for actions that are subject to binding by other automata are declared public.

@@

\paragraph*{Dynamic composition}
%% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
\ifjournal
The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a generator for allocating child automata.
Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class contain methods for destroying and unbinding respectively.
\ifjournal
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.
\fi

@@

\paragraph*{Handling errors by replacement}
The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.
To replace a child automaton, the parent automaton must destroy the child (thus dissolving all bindings with the child), create a new child, and bind to the new child.
Recall that system actions are asynchronous and can be executed in any order.
Thus, the user might might request the actions in one order, e.g., destroy, create, bind, but the system may execute them in a different order, e.g., create, bind, destroy.
If the sequence executed by the system automaton is (create, bind, destroy) then the bind will fail because the user automaton's actions are still bound to the old child.
Note that executing the bind action before the create will also fail but this case is prevent by the ioa::binding\_manager class.
Thus, a user automaton must receive notification that the first child has been destroyed before (creating and) binding to the second child.
A general solution is to observe the automaton and wait for it to be destroyed.

>>>>

All automata inherit from the common base class ioa::automaton which implements the system action interface.
\ifjournal
The following code declares the asynch\_lcr\_automaton:
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
The asynch\_lcr\_automaton is declared as a C++ template where the UID template parameter must model the \emph{unique identifier} concept.
The metaprogramming facilities of the C++ language aid in the development of general purpose reusable components.
\fi

@@

A number of wrappers and macros are provided by ioa++ for declaring actions.
%%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
Local actions are declared using three functions corresponding respectively to precondition, effect, and scheduling.
%%The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
\ifjournal
For example, the send action of the asynch\_lcr\_automaton is declared:
\begin{lstlisting}
private:
  bool send_precondition () const { ... }
  UID send_effect () { ... }
  void send_schedule () const { ... }
public:
  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
\end{lstlisting}
\fi
Input actions are declared similarly except for the absence of a precondition.
The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
The macros for actions that are subject to binding by other automata are declared public.

@@

\paragraph*{Dynamic composition}
%% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
\ifjournal
The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
%%Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a factory for allocating child automata.
%%Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class contain methods for destroying and unbinding respectively.
\ifjournal
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manage is destroyed or unbound respectively.
\fi

@@

\paragraph*{Handling errors by replacement}
The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.
To replace a child automaton, the parent automaton must destroy the child (thus dissolving all bindings with the child), create a new child, and bind to the new child.
Recall that system actions are asynchronous and can be executed in any order.
Thus, the user might might request the actions in one order, e.g., destroy, create, bind, but the system may execute them in a different order, e.g., create, bind, destroy.
If the sequence executed by the system automaton is (create, bind, destroy) then the bind will fail because the user automaton's actions are still bound to the old child.
Note that executing the bind action before the create will also fail but this case is prevented by the ioa::binding\_manager class.
Thus, a user automaton must receive notification that the first child has been destroyed before (creating and) binding to the second child.
A general solution is to observe the automaton and wait for it to be destroyed.

### 2011-09-30 papers/paper1/related_work.tex
refactored abstract and intro, updated my config
Another pass.

Our development of ioa++ grew out of the challenges faced when trying to develop distributed systems with threads.
Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
These same concerns are relevant to trends towards increasingly concurrent programming to increase performance, given the leveling trend of clock frequencies and the advent of multi-core processors\cite{sutter2005software}.

The direct implementation of a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
Structured programming simplifies development by allowing programmers to reason about their programs directly from the source code; regardless of whether of not the code it formally verified.
The I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} was selected because it supports features essential to developing distributed systems~\cite{lynch1996distributed}.
%% Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.

Incorporating a formal model of concurrency into a language to make concurrent programming easier is an established technique.
We consider three exemplars:  Communicating Sequential Processes, Actors, and threads.
The Communicating Sequential Processes (CSP) model~\cite{hoare1978communicating} is an extremely popular technique for modeling concurrent systems.
The fundamental concurrency control primitive in CSP is the \emph{rendezvous} which allows processes to synchronize and possibly exchange values.
CSP was influential in the design of the occam programming language~\cite{jones1987programming} and continues to influence programming language design, e.g., Go~\cite{go}.
An actor~\cite{agha1986actors} is a behavior that can perform a number of actions when receiving a message including (1) replacing itself with a new behavior, (2) sending messages to other actors, and (3) creating new actors.
The Actor model influenced the design of languages such as Erlang~\cite{armstrong1996concurrent} and Scala~\cite{odersky2004overview}.
The collective formalization of threads, e.g., semaphores~\cite{dijkstra1968cooperating}, monitors~\cite{hoare1974monitors}, all appear in popular libraries, e.g., POSIX threads~\cite{butenhof1997programming}, and languages, e.g., Java~\cite{christopher2000high}.

@@

The ioa++ framework can be compared directly to the IOA language due to the availability of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
\cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
Programs written in IOA are compiled to Java and executed on hosts communicating using the MPI library.
IOA requires programs to presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
IOA lacks support for procedure calls which complicates modeling and interfacing with external services: especially when a procedure call blocks.
The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
The next action to be executed is determined by a schedule function that resolves the non-determinism in the schedule.

IOA and ioa++ represent two fundamentally different approaches to programming directly with I/O automata.
The development of a new language was necessary for IOA due to their focus on formal verification.
On the other hand ioa++ uses the existing C++ language and compilers.
IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors, relying on the programmer to ensure non-blocking operation.
IOA focuses on static composition to produce a single node program while ioa++ focuses on dynamic composition and allows concurrent execution.
Both IOA and ioa++ require the programmer to add code to schedule actions, which may be a common source of errors.
Scheduling in IOA is deterministic and the programmer provides the policy.
Scheduling in ioa++ is non-deterministic according to the global policy encoded by the controller.

>>>>

Our development of ioa++ grew out of the challenges faced when trying to develop distributed systems with threads.
Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
Given the leveling of processor clock frequencies and the advent of multi-core, these same concerns are relevant to trends that look to deliver performance increases through increasingly concurrent programs~\cite{sutter2005software}.

The direct implementation of a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
Structured programming simplifies development by allowing programmers to reason about their programs directly from the source code; regardless of whether of not the code is formally verified.
The I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} was selected because it supports features essential to developing distributed systems~\cite{lynch1996distributed}.
%% Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.

Incorporating a formal model of concurrency into a language to make concurrent programming easier is an established technique.
We consider three exemplars:  Communicating Sequential Processes, Actors, and threads.
The Communicating Sequential Processes (CSP) model~\cite{hoare1978communicating} is an extremely popular technique for modeling concurrent systems.
The fundamental concurrency control primitive in CSP is the \emph{rendezvous} which allows processes to synchronize and possibly exchange values.
CSP was influential in the design of the occam programming language~\cite{jones1987programming} and continues to influence the design of modern programming languages, e.g., Go~\cite{go}.
An actor~\cite{agha1986actors} is a behavior that can perform a number of actions when receiving a message including (1) replacing itself with a new behavior, (2) sending messages to other actors, and (3) creating new actors.
The Actor model influenced the design of languages such as Erlang~\cite{armstrong1996concurrent} and Scala~\cite{odersky2004overview}.
The collective formalization of threads, i.e., semaphores~\cite{dijkstra1968cooperating}, monitors~\cite{hoare1974monitors}, etc., appears in popular libraries, e.g., POSIX threads~\cite{butenhof1997programming}, and languages, e.g., Java~\cite{christopher2000high}.

@@

The ioa++ framework can be compared directly to the IOA language due to the availability of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
\cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
Programs written in IOA are compiled to Java and executed on hosts communicating using the MPI library.
IOA requires programs to presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
IOA lacks support for procedure calls which complicates modeling and interfacing with external services: especially when a procedure call blocks.
The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
The next action to be executed is selected by a function that resolves the non-determinism in the schedule.

IOA and ioa++ represent two fundamentally different approaches to programming directly with I/O automata.
The development of a new language was necessary for IOA due to their focus on formal verification.
On the other hand ioa++ uses the existing C++ language and compilers.
IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors, relying on the programmer to ensure non-blocking operation.
IOA focuses on static composition to produce a single node program while ioa++ focuses on dynamic composition and allows concurrent execution.
Both IOA and ioa++ require the programmer to add code to schedule actions, which may be a common source of errors.
In IOA, the programmer provides a deterministic scheduling policy.
Scheduling in ioa++, however, is non-deterministic according to the global policy encoded by the controller.
